---
title: 'MSc Data Science: R for Big Data'
author: 'Pr. Charles Bouveyron (Email: charles.bouveyron@univ-cotedazur.fr / Twitter: @cbouveyron)'
date: Last updated on `r Sys.Date()`
output:
    html_document:
        toc: true
        toc_float:
          collapsed: false
          smooth_scroll: true
        toc_depth: 2
        number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

## Who I am?

A professor/researcher using R for more than 10 years and who developed more than 10 R packages.

## Organization of the course

The course will be organized as follows:

* lectures,
* pratical sessions,
* applied projects.

# Introduction

## The softwares for Data Science

R is one of the most used softwares for Data Science:

* <https://cran.r-project.org>
* <http://www.rstudio.com>  

But there are of course good alternatives:

* Python (the SciKitLearn initiative is very good and clearly is missing in R!)
* not free softwares: SAS, Matlab, ...


## Why R for Data Science?

R is very good for DS because of:

* connected to the researchers / newest technologies
* easy to pulish new codes
* a community behind the software

## Working with R - Lesson 101

To start a project in R: 

* you'll need a Rstudio session
* a blank script > open a new R script to do some serious project
* use only the console for simple calulations
* the plot panel allows to see graphical results

Two main commands:

* `run` is devoted to execute line by line the code
* `source` will execute the whole script at once

# R basics

> Note: the Base R Cheat Sheet can be downloaded on the Rstudio website: <https://www.rstudio.com/resources/cheatsheets/>

## R as simple calculus tool

R can be used to make simple (and less simple) caclulations:

```{r}
2+2
exp(-2)
```

All calculations are made using vectors (or matrices later) and the `[1]` is a trace of this vector nature


```{r}
rnorm(15)
```

This command produces a vector of 15 values and the number within brackets at the beginning of the lines indicate the rank of the values in the vector.


```{r}
x = 2
```

It is possible to store the results in variables. The `=` operator assigns the value to the variable. When all is ok, R does not display anything!

It is also possible to use the arrow operator `->` which usually works in the same way as the `=` operator

```{r}
a = c(1,2,3,4,5)
```

The `c()` function allows to create vectors. Warning: all vectors must be of the same type (numeric, characters, booleans)!

This is a vector of booleans: 

```{r}
a = c(TRUE,FALSE,TRUE,FALSE,FALSE)
```

All operators are working on vectors elements by elements:
```{r}
a = c(1,2,3,4,5)
b = c(2,1,4,6,1)
a / b
```

The resulting vector is of the same size of the two input vectors. 

> Note: the operator for matrix multiplication is %*%

```{r}
c = a / b
c^2
```

Here, the expentation is again element by element.

It is usually very easy to translate Maths into code:
```{r}
xbar = sum(a) / length(a)
xbar
```

## How to find a function in the documentation

The R documentation is very well structured and it is easy to find information from it. A search in the documentation is usually in two steps:

1. Search with a key word within the documentation:
```{r}
??student
```
2. Access the documentation of the wanted function:
```{r}
?t.test
```
The documentation is always organised in the same manner: 

* `Description`: a short description in English of the method implemented in the function
* `Usage`: the way the function should be used, with all input parameters and their default values
* `Arguments`: a detailed explenation of the input parameters
* `Values`: a detailed explenation of the output fields

There are plenty of already coded functions:

```{r}
mean(a)
sd(a)
range(a)
max(a)
```

> Note: missing values are represented in R with the `NA` type. It is possible to locate missing values using the `is.na()` function.

## Basics of ploting

In R, the generic plot function allows to output a graphic from some object:

```{r}
a = c(1,2,3,4,5)
b = c(2,1,4,6,1)
plot(b,a)
plot(b,a,pch=19,col='lightblue',cex=2,type='b')
title(main='My wonderful title')
```

> Note: a plot is build like a Lego, meaning it is not possible to mify something whitout restarting from zero!

## Loading and saving data

Two options:

1. Data in CSV file, or similar (mydata.csv)
2. Data in Rdata format (mydata.Rdata)

How to proceed for CSV:
```{r, eval=FALSE}
# Load the data into X
X = read.csv('path/toward/thefile.csv')

# Save some data into a file
write.csv(X,file='path/toward/thefile2.csv')
```

How to proceed for Rdata:
```{r, eval=FALSE}
# Load the data
load('path/toward/thefile.Rdata')

# Save some data into a file
save(X,Y,a,b,out,file='path/toward/thefile2.Rdata')
```

> Note: the `rm(list=ls())` allows to clean the working space.

## Exercice

Write a script allowing to load a vector file and "remove" the missing values.

```{r, eval=FALSE}
# Load data
x = read.csv("path/to/myfile.csv")
x = as.vector(x$x) # to ensure that x is a vector

# Count the number of NAs
prop = sum(is.na(x)) / lenght(x)

# If too many NAs, remove them
if (prop > 0.05){
  x = x[-which(is.na(x))]
}

# Otherwise, replace NAs with the mean
if (prop <= 0.05){
  m = mean(x,na.rm = TRUE)
  x[which(is.na(x))] = m
}
```


# Advanced R

# Programming with R

# Reporting with R


