---
title: 'MSc Data Science: R for Big Data'
author: 'Pr. Charles Bouveyron (Email: charles.bouveyron@univ-cotedazur.fr / Twitter:
  @cbouveyron)'
date: "Last updated on `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float:
          collapsed: true
          smooth_scroll: true
        toc_depth: 2
        number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

## Who I am?

A professor/researcher using R for more than 10 years and who developed more than 10 R packages.

## Organization of the course

The course will be organized as follows:

* lectures,
* pratical sessions,
* applied projects.

# Introduction

## The softwares for Data Science

R is one of the most used softwares for Data Science:

* <https://cran.r-project.org>
* <http://www.rstudio.com>  

But there are of course good alternatives:

* Python (the SciKitLearn initiative is very good and clearly is missing in R!)
* not free softwares: SAS, Matlab, ...


## Why R for Data Science?

R is very good for DS because of:

* connected to the researchers / newest technologies
* easy to pulish new codes
* a community behind the software

## Working with R - Lesson 101

To start a project in R: 

* you'll need a Rstudio session
* a blank script > open a new R script to do some serious project
* use only the console for simple calulations
* the plot panel allows to see graphical results

Two main commands:

* `run` is devoted to execute line by line the code
* `source` will execute the whole script at once

# R basics

> Note: the Base R Cheat Sheet can be downloaded on the Rstudio website: <https://www.rstudio.com/resources/cheatsheets/>

## R as simple calculus tool

R can be used to make simple (and less simple) caclulations:

```{r}
2+2
exp(-2)
```

All calculations are made using vectors (or matrices later) and the `[1]` is a trace of this vector nature


```{r}
rnorm(15)
```

This command produces a vector of 15 values and the number within brackets at the beginning of the lines indicate the rank of the values in the vector.


```{r}
x = 2
```

It is possible to store the results in variables. The `=` operator assigns the value to the variable. When all is ok, R does not display anything!

It is also possible to use the arrow operator `->` which usually works in the same way as the `=` operator

```{r}
a = c(1,2,3,4,5)
```

The `c()` function allows to create vectors. Warning: all vectors must be of the same type (numeric, characters, booleans)!

This is a vector of booleans: 

```{r}
a = c(TRUE,FALSE,TRUE,FALSE,FALSE)
```

All operators are working on vectors elements by elements:
```{r}
a = c(1,2,3,4,5)
b = c(2,1,4,6,1)
a / b
```

The resulting vector is of the same size of the two input vectors. 

> Note: the operator for matrix multiplication is %*%

```{r}
c = a / b
c^2
```

Here, the expentation is again element by element.

It is usually very easy to translate Maths into code:
```{r}
xbar = sum(a) / length(a)
xbar
```

## How to find a function in the documentation

The R documentation is very well structured and it is easy to find information from it. A search in the documentation is usually in two steps:

1. Search with a key word within the documentation:
```{r}
??student
```
2. Access the documentation of the wanted function:
```{r}
?t.test
```
The documentation is always organised in the same manner: 

* `Description`: a short description in English of the method implemented in the function
* `Usage`: the way the function should be used, with all input parameters and their default values
* `Arguments`: a detailed explenation of the input parameters
* `Values`: a detailed explenation of the output fields

There are plenty of already coded functions:

```{r}
mean(a)
sd(a)
range(a)
max(a)
```

> Note: missing values are represented in R with the `NA` type. It is possible to locate missing values using the `is.na()` function.

## Basics of ploting

In R, the generic plot function allows to output a graphic from some object:

```{r}
a = c(1,2,3,4,5)
b = c(2,1,4,6,1)
plot(b,a)
plot(b,a,pch=19,col='lightblue',cex=2,type='b')
title(main='My wonderful title')
```

> Note: a plot is build like a Lego, meaning it is not possible to mify something whitout restarting from zero!

## Loading and saving data

Two options:

1. Data in CSV file, or similar (mydata.csv)
2. Data in Rdata format (mydata.Rdata)

How to proceed for CSV:
```{r, eval=FALSE}
# Load the data into X
X = read.csv('path/toward/thefile.csv')

# Save some data into a file
write.csv(X,file='path/toward/thefile2.csv')
```

How to proceed for Rdata:
```{r, eval=FALSE}
# Load the data
load('path/toward/thefile.Rdata')

# Save some data into a file
save(X,Y,a,b,out,file='path/toward/thefile2.Rdata')
```

> Note: the `rm(list=ls())` allows to clean the working space.

## Exercice

Write a script allowing to load a vector file and "remove" the missing values.

```{r, eval=FALSE}
# Load data
x = read.csv("path/to/myfile.csv")
x = as.vector(x$x) # to ensure that x is a vector

# Count the number of NAs
prop = sum(is.na(x)) / lenght(x)

# If too many NAs, remove them
if (prop > 0.05){
  x = x[-which(is.na(x))]
}

# Otherwise, replace NAs with the mean
if (prop <= 0.05){
  m = mean(x,na.rm = TRUE)
  x[which(is.na(x))] = m
}
```


# Advanced R

## Objects and expressions

First, it important to understand that R has two kinds of "objects":

* expresssions: operators and functions,
* objects: vectors, matrices, lists, dataframes, ...

To differentiate between functions and objects, functions have to be called every time with brackets `()`!

```{r}
plot = c(1,1,2,3,4,5)
plot(plot)
```

About the functions, the arguments (imput parameters) can be provided in any order:

```{r}
x = c(1,2,3,4,5,6)
y = sin(x)
plot(x,y,type='b')
```

or 
```{r}
plot(y=y,x=x,col='red',type='b')
```
if you declare clearly which argument you are providing!

## Vectors

Vectors can be created with the `c()` function:

```{r}
x = c(1,2,3,4,5)
x = c(TRUE,FALSE,TRUE)
x = c('tyty',"tutu",'yuyu')
```

or some other ways:

```{r}
x = seq(1,10)
x
```

```{r}
x = seq(1,11,by = 1.5)
x
```

```{r}
x = seq(1,10,length.out = 100)
x
```

This would be very useful to get a nice plot of the `sinus` function.

```{r}
x = seq(-pi,pi,length.out = 100)
plot(x,sin(x),type='l')
```

You can also use the `:` operator which equivalent to the default `seq()` function.

```{r}
x = seq(1:10)
x

y = 1:10
y
```

```{r}
w = 0.5:10
w
```

And the last way of creating vectors is to use the `rep` function.

```{r}
x = rep(0,10)
x
```

This is useful in particular to initialize a vector which will be used in a loop or a function.

> Tip: always initialize your vectors with a "strange" value for the task (for instance NA's)

```{r}
m = rep(NA,10)
for (i in 1:10) m[i] = mean(rnorm(10))
m
```


## Matrices

**Matrices** can be created with the `matrix` function:

```{r}
A = matrix(c(1,2,3,4,5,6),nrow = 2,byrow = TRUE)
A
```

Another option would be to use the `cbind` and `rdind` functions:

```{r}
A = cbind(c(1,2),c(2,3),c(4,5))
A
```



```{r}
A = rbind(c(1,2,3),c(4,5,6))
A
```

To access vectors and matrices, we cannot use the barckets ()! Instead, we use the square brackets `[...]`

```{r}
x = seq(1,10,length.out = 25)
# Access the 5th element of x
x[5]
# Access all elements between the 5th and the 15th elements
x[5:15]
# Access all elements except those between 5 and 15
y = x[-(5:15)]
y
length(y)
```

With the square brackets, you can also modify the values of a vector:

```{r}
x[5] = NA
x
```

It is also possible to do some conditional access:
```{r}
x[is.na(x)] = -1
x
```

```{r}
x[x < 4] = 0
x
```

```{r}
x[x < 4 | x > 8] = NA
x
```

> Note: the `|` operator denotes a OR. The AND is denoted by the `&` operator.

For matrices, you will have to provide position information about rows and columns:

```{r}
A = diag(1,5)
A
```

```{r}
A[1:2,4:5]
```

If you would like to access all rows (resp. columns) of a matrix, you can use the "nothing" operator:

```{r}
A[,4:5]
```

In some cases, you perhaps would like to access some specific locations:
```{r}
A[c(1,3),c(2,5)]
```

## Lists

**Lists** are objects able to store information of different types. All functions return lists, so it is important to be able to manipulate them!

```{r}
mylist = list(name='Charles',age=28,married=TRUE)
mylist
```

```{r}
mylist$name = 'Tutu'
mylist
```

```{r}
mylist$Nb_of_cats = 4
mylist
```

```{r}
out = t.test(1:10, y = c(7:20))
out
```

```{r}
print.default(out)
```


```{r}
out$statistic
sqrt(abs(out$statistic))
```

```{r}
out$p.value
```

## Data frames

The last object is the `data.frame` and it is devoted to store data sets. In a `data.frame`, rows are for individuals and columns are for variables.

```{r}
# We create a data.frame with 28 individuals and 3 variables (2 numeric and 1 categorical)
X = data.frame(x = seq(1,10,length.out = 28),
               y = seq(-3,3,length.out = 28),
               sex = c(rep('M',10),rep('F',18)))
X
```

Data.frame can be used as matrices or lists:
```{r}
X[5:10,]
```

```{r}
X$sex
```

```{r}
X[,3]
```

```{r}
summary(X)
```

Some useful functions to get information about objects:
```{r}
#Length of a vector
length(x)

# Length of a list (nb of fields)
length(mylist)
```

```{r}
# For matrices and data.frames
nrow(X)

ncol(X)
```

```{r}
# For data.frames
colnames(X)
rownames(X)
```

It is also possible to modify the names of variables and individuals

```{r}
colnames(X) = c('Var 1','Var 2','Var 3')
X
```

```{r}
rownames(X) = paste('Individual',1:28)
X
```

```{r}
X["Individual 26",]
```

## Some exercises

> Exercise: Create two vectors of different dimensions and insert the second one in the first one between the 2nd and 3rd elements.

```{r}
a = c(1,2,3,4,5)
b = runif(10)
c = c(a[1:2],b,a[-(1:2)])
c
```

> Exercise: Draw 100 numbers from a Uniform distribution on [0,1] and count how many values are larger than 0.5

```{r}
x = runif(100,min = 0,max = 1)
sum(x>=0.5)
length(x[x>=0.5])
```

> Exercise: Draw 100 couples $(x,y)$ randomly in the square $[0,1]^2$ (use a data.frame to store the data). Count how many couples are within the square $[0.5,1]^2$. Plot the couples and display with a red point the mean of all points.

```{r}
X = data.frame(x = runif(100), y = runif(100))
sum(X$x >= 0.5 & X$y >= 0.5)
plot(X,col='lightblue',pch=19)
mean = colMeans(X)
points(mean[1],mean[2],col='red',pch=19,cex=2)
```


> Exercise: It is Spring! We consider a field of flower bulbes. We have 100 bulbes in a field of 1m x 1m. We know that the probability that a bulbe blooms is 0.85 (use a data.frame to store the data). Simulate the blossoming of the bulbes!

```{r}
X = data.frame(x = runif(100), y = runif(100),
               b = rbinom(100,size = 1,prob = 0.85))
X
plot(X[,1:2],pch=19,col=X$b+1)
nb = sum(X$b == 1)
title(main = paste(nb,"bulbes become flowers!"))
legend("bottomleft",legend = c('Failed','Bloomed'),col=c(1,2),pch=19)
```

# The graphical system

## Basic plots and legends

The graphical system is very complex and works in Lego fashion, i.e. it is not possible to remove something on plot wihout clearing everything!

Of course, it is possible to add on a figure texts, legend, title, etc ... and to save the figure in an image format (png, jpg, pdf, eps, ...)

The window of the plot is automatically set up according to the data scale.
```{r}
x = runif(100)
y = rnorm(100)
plot(x,y,pch=19)
```

It is possible to resize the window by using the `xlim` and `ylim` arguments.
```{r}
plot(x,y,pch=19,xlim=c(-1,2),ylim=c(-3,4))
points(1.5,0,col='red')
points(0,3,col='blue')
```

```{r}
plot(x[x >= 0.5],y[x >= 0.5],pch=19,
     xlab='x',ylab='y',main='My title')
title(sub='My subtitle')
legend("bottomright",legend = 'my points',col=1,pch=19)
text(x,y,1:length(x >= 0.5),pos = 1)
```

To draw some line (for instance for linear regression), we use the `abline` function:

```{r}
plot(x,x+rnorm(100,0,0.2),pch=19,ylab='y')
abline(a=0,b=1,lty=2,lwd=2,col='red')
abline(v = 0.5,lty=3,col='blue')
```

It is also possible to add texts in the margin:

```{r}
plot(x,x+rnorm(100,0,0.2),pch=19,ylab='y')
abline(a=0,b=1,lty=2,lwd=2,col='red')
abline(v = 0.5,lty=3,col='blue')
mtext("A nice linear model",side = 4)
```

## Basic statistics with R plots

Histograms are a way to look at the distribution of the data. 

```{r}
x = rnorm(200,mean = 2,sd = 1.5)
hist(x,col='lavender')
```

Be careful with histograms: it is not so easy to draw a great histogram! In addition, keep in mind that it is possible to manipulate a histogram to pass a specific message!

For example, this is the default result for a uniformly distributed variable:

```{r}
x = runif(100,0,1)
hist(x)
```

And this are other possible histograms for the same data:

```{r}
par(mfrow=c(1,2))
hist(x,breaks=2)
hist(x,breaks=100)
```

It is possible to combine or replace the histogram with a plot o the estimated density function. The function to estimate the density function is `density`:

```{r}
x = c(rnorm(100,2,1),rgamma(50,shape = 1))
hist(x,freq = FALSE)
f = density(x)
lines(f,lwd=2,col='red',lty=2)
```

It is possible to compare the empirical distribution of the data with the theorical function of some specific distribution:

```{r}
x = c(rnorm(100,2,1),rgamma(50,shape = 1))
qqnorm(x)
```

```{r}
x = c(rnorm(100,2,1),rgamma(50,shape = 1))
qqplot(x,y = rgamma(1000,shape=1))
```

Perhaps, the best way to look at the distribution of some data is to do a boxplot. Boxplots are great becaus the recipie is unique!

```{r}
par(mfrow=c(1,2))
boxplot(rnorm(100),title='Normal')
boxplot(runif(100),title='Uniform')
```


```{r}
par(mfrow=c(1,2))
boxplot(rnorm(100),title='Normal')
boxplot(rgamma(100,shape=1),title='Gamma')
```

But, it is a very simple plot and it is not easy to look at complex distributions:

```{r}
x = c(rnorm(100,2,1),rgamma(50,shape = 1))
boxplot(x,title='Mixture dist.')
```

Interestingly, it is possible to display multavariate data with boxplots:

```{r}
x = iris[,-5]
boxplot(x)
```

This is very helpful but limited to data on the same unit! 

To represent multivariate data, we also have the `pairplot` funtion.

```{r}
x = iris[,-5]
pairs(x)
```

> Remark: Keep in mind that some information can remain hidden in high-dimensional spaces!

It is possible to add categorical variables on a plot by changing the colors or the forms of the points.

```{r}
x = iris[,-5]
y = iris$Species
pairs(x,col=as.numeric(y),pch=as.numeric(y))
```

Finally, for categorical variables, you can draw pie plots:

```{r}
y = iris$Species
pie(summary(y))
```

## The ggplot package

We are going to use an extra package, `ggplot`, that can be download on the CRAN repository.

```{r eval=FALSE, include=FALSE}
#install.packages('ggplot2')
library(ggplot2)
```

We are going to use the `mpg` data set to illustrate the use of the package:

```{r}
library(ggplot2)
?mpg
mpg
```


The ggplot package has a very specific wayt of working:
```{r}
ggplot(data = mpg) + geom_histogram(aes(x = hwy))
```

> Notice that `geom_histogram` always use 30 bins for making the plot! It is necessary to provide a better value if you want!

```{r}
ggplot(data = mpg) + geom_histogram(aes(x = hwy), bins=50, fill = 'pink')
```

```{r}
ggplot(data = mpg) + geom_histogram(aes(x = hwy, y = ..density..), bins=50, fill = 'pink') + geom_density(aes(x = hwy),col = 'green')
```

It is also possible to draw boxplots:

```{r}
boxplot(mpg$hwy)
ggplot(data = mpg) + geom_boxplot(aes(x = '',y = hwy), fill = 'lightblue')
```

Let's compare the distribution of the `hwy` variable with one normal distribution with the quantile-quantile plot (qqplot for friends!):

```{r}
ggplot(data = mpg) +
  geom_qq(aes(sample = hwy)) +
  geom_qq_line(aes(sample = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_qq(aes(sample = hwy),distribution = stats::qgamma,dparams = (shape=1)) + 
    geom_qq_line(aes(sample = hwy),distribution = stats::qgamma,dparams = (shape=1))
```

For categorical data, we also have some barplots:

```{r}
ggplot(data = mpg) + geom_bar(aes(x = trans))
```

It is also possible to have some original graphics that are not easy to obtain with the usual plot function, like this one:

```{r}
ggplot(data = mpg) + geom_bar(aes(x = "", fill = factor(trans)))
```

To draw a barplot:

```{r}
ggplot(data = mpg) + geom_bar(aes(x = "", fill = factor(trans))) + coord_polar(theta = "y")
```

For two-dimensional data, we have also specific plots:

```{r}
ggplot(data = mpg) + geom_point(aes(x = displ,y = hwy))
```

```{r}
ggplot(data = mpg) + geom_point(aes(x = displ,y = hwy,col = class, shape = factor(year)))
```

It is also possible to display the data conditionnaly on one categorical variable (here the `class` variable):

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, col = class)) + facet_wrap( ~ class, nrow = 3)
```

It is even possible to make the plots conditionnaly on two categorical variables:

```{r}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy)) + facet_grid(drv ~ cyl)
```

Let's try to do the same without ggplot:

```{r}
par(mfrow=c(3,4))
for (m in unique(mpg$drv))
  for (v in unique(mpg$cyl)){
    sel = which(mpg$drv == m & mpg$cyl == v)
    if (length(sel)>0) plot(mpg$displ[sel],mpg$hwy[sel],pch=19)
    else plot(0,0,type='n')
  }
```


```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy),method = 'lm')
```

If one wuld like to interpret the coefficients of the `lm` model, it is necessary to use directrly the lm function:

```{r}
out = lm(mpg$hwy ~ 1 + mpg$displ)
out
```

```{r}
plot(out)
```

```{r}
plot( mpg$displ, mpg$hwy, type='p')
abline(out, col='red')
```

It is worth noticing that the ggplot graphic is more meaningful regarding the distribution of the residuals.

When working with multivariate categorical data, it still possible to have nice plots!

```{r}
table(mpg$cyl,mpg$drv)
```

```{r}
ggplot(data = mpg) + 
  geom_bar(aes(x = factor(cyl), fill = factor(drv)))
```

```{r}
ggplot(data = mpg) + 
  geom_bar(aes(x = factor(drv), fill = factor(cyl)))
```

There are variants of this plot, that can be obtained using the `position` option.


```{r}
ggplot(data = mpg) + geom_bar(aes(x = factor(cyl), fill = factor(drv)), position = "dodge")

ggplot(data = mpg) + geom_bar(aes(x = factor(cyl), fill = factor(drv)), position = "fill")

```

```{r}
ggplot(data = mpg) + 
  geom_bar(aes(x = "", fill = factor(cyl)), position = "fill") + 
  coord_polar(theta = "y") +
  facet_wrap(~ factor(drv))
```

> Remark: in the two last plots, information about the counts are missing and this could be dangerous when intyerpreting... 

```{r}
tapply(mpg$displ, mpg$cyl, mean)

ggplot(data = mpg) + geom_boxplot(aes(x = factor(cyl), y = displ))

ggplot(data = mpg) + geom_histogram(aes(x = displ)) + facet_wrap(~ factor(cyl), ncol = 1)

ggplot(data = mpg) + geom_density(aes(x = displ, col = factor(cyl)))
```


> Exercise: Use the `starwars data set in the `dplyr` package to:

  - list the different human characters,
  - list the different worlds,
  - compute the average weight and height of the different character types,
  - display on a plot the number of characters of each type in a deacresing order,
  - visualize the relationship between the height and weight of the different characters.

```{r}
library(dplyr)
?starwars
starwars$name[starwars$species == 'Human']
```
```{r}
unique(starwars$homeworld)
#levels(as.factor(starwars$homeworld))
```

```{r}
weight = tapply(starwars$mass, starwars$species, mean)
height = tapply(starwars$height, starwars$species, mean)
Species = data.frame(weight,height)
Species
ggplot(data = Species) + geom_point(aes(x = weight,y = height)) +
  geom_text(aes(x = weight,y = height,label = rownames(Species)),nudge_y = 10)
```

```{r}
Species$count = table(starwars$species)
Species = Species[order(Species$count,decreasing = TRUE),]
#ggplot(data = Species) + geom_bar(aes(x = count))
barplot(Species$count)
```

```{r}
ggplot(data = starwars) +
  geom_point(aes(x = mass,y = height)) +
  geom_smooth(aes(x = mass,y = height),method = 'lm')
```

> Remark: we can see here that the linear model is very sensitive to ouliers!  


# Reporting with R

Several very useful tools for reporting and communication are available within the R ecosystem. 

## Rmarkdown

Rmarkdown is a comprehensive way of creating reports which involve R (or other languages). A synthetic presentation of the Rmarkdown system is available here: <https://rmarkdown.rstudio.com/lesson-1.html>. For advanced details on the language, please refer to <https://bookdown.org/yihui/rmarkdown/>.

## Shiny

# Programming with R

## Basics of R programming

## Advanced tools

## Building R packages


# Handling massive data sets with `dplyr`